---
title: "glowing_wave"
format: html
---

# Monterey Bay beaked whale seasonal presence

```{r}
#| label: setup

library(tidyverse)
library(lubridate)
library(cowplot)
library(mgcv)
source(here::here("R/bw_funcs.R"))

zc_mb <- read_csv(here::here("data/processed_bw/zc_monterey.csv")) 

zc_mb_monthly <- zc_mb %>% 
  mutate(date = ymd(Date),
         month = Month,
         year = year(Date)) %>%
  arrange(date) %>%
  #this is okay because I know we had full effort in 2016-18, but will need to change if adding data from years where we had off effort months
  complete(year = 2016:2018, month = 1:12, fill = list(count = 0)) %>%
  distinct(year, month, count) %>%
  mutate(date = ISOdate(year, month, day = 1))

#daily effort at MARS
load(here::here("data/MARS_DailyPercentRecorded.RData")) 
P <- P %>% 
  rename(Date = daystart) %>% 
  filter(year(Date) < 2019)

#daily presence, filtered to days when effort was > 85% (not sure this is the right way to do this)
zc_mb_daily <- P %>% 
  left_join(
    zc_mb %>% 
      distinct(Date) %>% 
      mutate(present = TRUE), 
    by = "Date"
  ) %>% 
  mutate(present = replace_na(present, FALSE),
         month = month(Date), 
         year = factor(year(Date))) %>% 
  filter(PercentRecorded >= 85)

ggplot(zc_mb_daily, aes(Date, present)) + 
  geom_point()


```

### Summary stats

```{r}
zc_mb %>% 
  summarize(across(count, list(mean = mean, sd = sd, total = length))) 
```

### Plot encounters for each species in Monterey Bay.

```{r}

zc_raw_plot <- ggplot(zc_mb_monthly, aes(month, count)) +
  geom_area(aes(fill = factor(year))) +
  scale_fill_manual(values = c("orange", "tomato", "firebrick")) + 
  labs(x = "Month", 
       y = "Zc (days)") +
  scale_x_continuous(breaks = 1:12, minor_breaks = NULL) +
  theme_bw() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.01, 0.99),
        legend.justification = c(0, 1),
        legend.title = element_blank(),
        legend.background = element_rect(fill = NA))

```

### Fit GAMs for seasonal presence

Zc:

```{r}
zc_mb_daily_mod <- gam(present ~ 
                         s(month, bs = "cc", k = 12) +
                         s(year, bs = "re"),
                       family = binomial(link = "logit"),
                       data = zc_mb_daily,
                       method = "REML")

summary(zc_mb_daily_mod)
gam.check(zc_mb_daily_mod)
acf(residuals(zc_mb_daily_mod))

zc_mb_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(zc_mb_daily_mod,
                newdata = zc_mb_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

zc_mb_grid <- zc_mb_grid %>%
  mutate(
    fit = plogis(pred$fit),
    lwr = plogis(pred$fit - 1.96 * pred$se.fit),
    upr = plogis(pred$fit + 1.96 * pred$se.fit)
  ) %>% 
  mutate(across(fit:upr, \(x) x / pracma::trapz(month, fit), .names = "{.col}_norm"))

zc_mb_peaks <- find_peaks(zc_mb_grid)
zc_mb_peaks$fit_norm <- zc_mb_peaks$fit / pracma::trapz(zc_mb_grid$month, zc_mb_grid$fit)

# Clip grid back to 1-12 for plotting
zc_mb_norm_plot <- ggplot(zc_mb_grid %>% filter(month >= 1, month <= 12), 
                          aes(x = month, y = fit_norm)) +
  geom_ribbon(aes(ymin = lwr_norm, ymax = upr_norm), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(data = zc_mb_peaks[2:3,], aes(x = month, y = fit_norm),
             size = 3, shape = 21, fill = "white", stroke = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  # ylim(0, .2) + 
  theme_bw() +
  labs(x = NULL, 
       y = NULL)

zc_mb_plot <- ggplot(zc_mb_grid %>% filter(month >= 1, month <= 12), 
                          aes(x = month, y = fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(data = zc_mb_peaks[2:3,], aes(x = month, y = fit),
             size = 3, shape = 21, fill = "white", stroke = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  # ylim(0, .2) + 
  theme_bw() +
  labs(x = NULL, 
       y = "Zc")

```

Both plots together:

```{r}

plot_grid(zc_raw_plot, zc_mb_plot, 
          nrow = 2, 
          align = "h")

ggsave("figs/mb_zc.svg", width = 4, height = 4)
```

### BEUTI in Monterey Bay

```{r}

beuti <- read_csv(here::here("data/indices/BEUTI_06-25.csv")) %>% 
  set_names(c("date", "latitude", "beuti")) %>% 
  filter(latitude == 37) %>% 
  mutate(month = month(date), 
         year = year(date))

#climatology to get at the long term average
beuti_clim_mod <- gam(beuti ~ 
                        s(month, bs = "cc", k = 12) +
                        s(year, bs = "re"),
                      family = gaussian(),
                      data = beuti,
                      method = "REML")

mb_beuti_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(beuti_clim_mod,
                newdata = mb_beuti_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

mb_beuti_grid$beuti <- pred$fit
mb_beuti_grid$lower <- pred$fit - pred$se.fit * 1.96
mb_beuti_grid$upper <- pred$fit + pred$se.fit * 1.96

mb_beuti_grid %>% 
  filter(between(month, 1, 12)) %>% 
  ggplot(aes(month, beuti)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line() + 
  theme_classic() +
  scale_x_continuous(breaks = 1:12) + 
  geom_vline(xintercept = zc_mb_peaks[2:3,]$month, 
             color = "tomato", 
             linetype = "dashed") + 
  ylim(0, 20) + 
  ggtitle("Monterey Bay")

ggsave("figs/mb_beuti.svg", width = 2.8, height = 2.5)

# Peak lasts 1.5 months
sum(mb_beuti_grid$beuti > 0.9 * max(mb_beuti_grid$beuti)) * 0.1
mb_beuti_grid$month[which.max(mb_beuti_grid$beuti)]

```

### Curve simulation

Simulating 10,000 seasonal curves for each species of beaked whale and for BEUTI.

```{r}
n_sim <- 1e4
set.seed(123)
```

Zc:

```{r}
### Zc ###

Xp_zc <- predict(zc_mb_daily_mod, 
                 newdata = zc_mb_grid, 
                 type = "lpmatrix")
coef_sims_zc <- MASS::mvrnorm(n_sim, 
                              mu = coef(zc_mb_daily_mod), 
                              Sigma = vcov(zc_mb_daily_mod))
sim_curves_zc <- Xp_zc %*% t(coef_sims_zc)
sim_curves_response_zc <- zc_mb_daily_mod$family$linkinv(sim_curves_zc)
boot_zc <- expand_grid(i = 1:nrow(zc_mb_grid), 
            j = 1:ncol(sim_curves_response_zc)) %>% 
  mutate(month = zc_mb_grid$month[i], 
         sim = j, 
         count = sim_curves_response_zc[cbind(i, j)]) %>% 
  select(-c(i, j)) 
#warning is because there is a year = 0 placeholder in the prediction grid; ok since exclusing the year smooth

### BEUTI ###
Xp_beuti <- predict(beuti_clim_mod, newdata = mb_beuti_grid, type = "lpmatrix")
coef_sims_beuti <- MASS::mvrnorm(n_sim, mu = coef(beuti_clim_mod), Sigma = vcov(beuti_clim_mod))
sim_curves_beuti <- Xp_beuti %*% t(coef_sims_beuti)
```

### Calculating lags for each species.

Zc:

```{r}
boot_beuti <- expand_grid(i = 1:nrow(mb_beuti_grid), 
            j = 1:ncol(sim_curves_beuti)) %>% 
  mutate(month = mb_beuti_grid$month[i], 
         sim = j, 
         beuti = sim_curves_beuti[cbind(i, j)]) %>% 
  select(-c(i, j)) 

boot_beuti_peak <- boot_beuti %>% 
  group_by(sim) %>% 
  summarize(peak_month_beuti = month[which.max(beuti)])

boot_zc_peak <- boot_zc %>% 
  group_by(sim) %>% 
  summarize(peak_month_zc = month[which.max(count)])

boot_peak_lag_zc <- left_join(boot_beuti_peak, boot_zc_peak, by = "sim") %>% 
  mutate(peak_lag = peak_month_zc - peak_month_beuti, 
         peak_lag = ifelse(peak_lag <= 0, peak_lag + 12, peak_lag)) 

mean_peak_lag_zc <- CircStats::circ.mean(boot_peak_lag_zc$peak_lag * 2 * pi / 12) * 12 / 2 / pi
ci_peak_lag_zc <- shortest_arc_ci(boot_peak_lag_zc$peak_lag)

if (mean_peak_lag_zc < 0) mean_peak_lag_zc <- mean_peak_lag_zc + 12

ggplot(boot_peak_lag_zc, aes(peak_lag)) + 
  geom_histogram() + 
  geom_vline(xintercept = mean_peak_lag_zc, color = "red") + 
  ggtitle("Zc") + 
  theme_classic()

ggsave("figs/mb_lag.svg", width = 2.8, height = 2.5)

boot_peak_lag_zc %>% 
  mutate(group = case_when(
    between(peak_lag, 6, 8) ~ "primary peak",
    peak_lag > 11 | peak_lag < 1 ~ "secondary peak",
    .default = "non peak"
  )) %>% 
  group_by(group) %>% 
  summarize(N = n()) %>% 
  mutate(prop = N / sum(N))

#in 52% of bootstrapped samples, the lag occured between months 6 and 8. In 27% of samples, the lag was approximately 0 (greater than 11 or less than 1). 
```

### Circular plot

```{r}
# normalize each variable to 0-1
zc_plot <- zc_mb_monthly %>% 
  group_by(month) %>%
  summarize(count = mean(count)) %>% 
  mutate(value = (count - min(count)) / (max(count) - min(count)))

beuti_clim <- beuti %>% 
  group_by(latitude, month) %>% 
  summarize(beuti = mean(beuti))

beuti_plot <- beuti_clim %>% 
  mutate(value = (beuti - min(beuti)) / (max(beuti) - min(beuti)))

brooding <- data.frame(
  month = 1:12,
  brooding = c(1, 1, 1, 1, 1, 1, 0, 0, 0, 1, 1, 1)  # Jan-Jun, Oct-Dec
)

ggplot() +
  geom_rect(data = brooding %>% filter(brooding == 1),
            aes(xmin = month - .5, xmax = month + .5,
                ymin = 0, ymax = 1),
            fill = "darkred", alpha = 0.2,
            inherit.aes = FALSE) +
  geom_col(data = zc_plot,
           aes(x = month, y = value),
           fill = "darkorange",
           alpha = 0.8,
           width = 0.9) +
  geom_line(data = beuti_plot, 
            aes(x = month, y = value), 
            color = "steelblue", 
            linewidth = 1) +
  scale_x_continuous(breaks = 1:12, 
                     labels = month.abb) +
  coord_polar(start = -pi/12) +
  ylim(0, 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(color = NULL)

ggsave("figs/mb_circle.svg", width = 2.8, height = 2.5)

```

# Zc presence across CA Current

### SCB GAM

```{r}

scb_zc <- read.csv(here::here("data/processed_bw/zc_scb.csv")) 
scb_zc$site <- factor(scb_zc$site)
scb_zc <- filter(scb_zc, site == "H")

scb_zc_raw_plot <- ggplot(scb_zc, aes(factor(month), Zc_pres_eff_adj)) +
  geom_boxplot() +
  labs(x = "Month", 
       y = "Zc presence (mins)") +
  scale_x_discrete(limits = factor(1:12), labels = month.abb) +
  theme_bw()

scb_bw_mod <- gam(Zc_pres_eff_adj ~ 
                    s(month, bs = "cc", k = 12) +
                    s(year, bs = "re"),
                  family = tw(link = "log"),
                  data = scb_zc,
                  method = "fREML",
                  knots = list(month = c(1, 12)))

scb_bw_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  site  = unique(scb_zc$site),
  year  = 0
) %>% mutate(site = factor(site, levels = levels(scb_zc$site)))

pred <- predict(scb_bw_mod,
                newdata = scb_bw_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

scb_bw_grid <- scb_bw_grid %>%
  mutate(
    fit = exp(pred$fit),
    lwr = exp(pred$fit - 1.96 * pred$se.fit),
    upr = exp(pred$fit + 1.96 * pred$se.fit)
  ) %>% 
  mutate(across(fit:upr, \(x) x / pracma::trapz(month, fit), .names = "{.col}_norm"))

# Find peaks on extended grid, then fold month back to 1-12
scb_peaks <- scb_bw_grid %>%
  group_by(site) %>%
  group_modify(~ find_peaks(.x)) %>%
  ungroup()

scb_peaks$fit_norm <- scb_peaks$fit / pracma::trapz(scb_bw_grid$month, scb_bw_grid$fit)

# Clip grid back to 1-12 for plotting
scb_norm_plot <- ggplot(scb_bw_grid %>% filter(month >= 1, month <= 12), 
       aes(x = month, y = fit_norm)) +
  geom_ribbon(aes(ymin = lwr_norm, ymax = upr_norm), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(data = scb_peaks, aes(x = month, y = fit_norm),
             size = 3, shape = 21, fill = "white", stroke = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_bw() +
  theme(legend.position = "none") + 
  labs(x = NULL, 
       y = NULL)
```

### SRI GAM

```{r}
sri_zc <- read_csv(here::here("data/processed_bw/zc_sri.csv"))
sri_zc_daily <- read_csv("data/processed_bw/zc_sri_daily.csv")

sri_zc_raw_plot <- ggplot(sri_zc, aes(factor(month), mins_eff_adj)) +
  geom_col(aes(fill = factor(year))) +
  labs(x = "Month", 
       y = "Zc presence (mins)") +
  scale_x_discrete(limits = factor(1:12), labels = month.abb) +
  theme_bw()

#FIX THIS BACK TO CLICK POS MINUTES
sri_mod <- gam(mins_eff_adj ~ 
                 s(month, bs = "cc", k = 12) +
                 s(year, bs = "re"),
               family = tw(link = "log"),
               data = sri_zc,
               method = "REML")

sri_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(sri_mod,
                newdata = sri_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

sri_grid <- sri_grid %>%
  mutate(
    fit = exp(pred$fit),
    lwr = exp(pred$fit - 1.96 * pred$se.fit),
    upr = exp(pred$fit + 1.96 * pred$se.fit)
  ) %>% 
  mutate(across(fit:upr, \(x) x / pracma::trapz(month, fit), .names = "{.col}_norm"))

sri_peaks <-  find_peaks(sri_grid)
sri_peaks$fit_norm <- sri_peaks$fit / pracma::trapz(sri_grid$month, sri_grid$fit)

# Clip grid back to 1-12 for plotting
sri_norm_plot <- ggplot(sri_grid %>% filter(month >= 1, month <= 12), 
       aes(x = month, y = fit_norm)) +
   geom_ribbon(aes(ymin = lwr_norm, ymax = upr_norm), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(data = sri_peaks[1,], aes(x = month, y = fit_norm),
             size = 3, shape = 21, fill = "white", stroke = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_bw() +
  labs(x = NULL, 
       y = NULL)
```

### OR GAM

```{r}
or_zc_monthly <- read.csv(here::here("data/processed_bw/zc_or.csv"))
or_zc <- read.csv(here::here("data/processed_bw/zc_or_daily.csv"))
or_zc <- or_zc %>% 
  group_by(month, year) %>% 
  mutate(count = sum(present))

or_raw_plot <- ggplot(or_zc, aes(month, count)) +
  geom_area(aes(fill = factor(year))) +
  scale_fill_manual(values = c("deeppink4", "violetred2")) + 
  labs(x = "Month", 
       y = "Zc (days)") +
  scale_x_continuous(breaks = 1:12, minor_breaks = NULL) +
  theme_bw() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.01, 0.99),
        legend.justification = c(0, 1),
        legend.title = element_blank(),
        legend.background = element_rect(fill = NA))

or_mod <- gam(present ~ 
                    s(month, bs = "cc", k = 10) + 
                    s(year, bs = "re"),
                  family = binomial(link = "logit"),
                  data = or_zc,
                  method = "REML")


or_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(or_mod,
                newdata = or_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

or_grid <- or_grid %>%
  mutate(
    fit = plogis(pred$fit),
    lwr = plogis(pred$fit - 1.96 * pred$se.fit),
    upr = plogis(pred$fit + 1.96 * pred$se.fit)
  ) %>% 
  mutate(across(fit:upr, \(x) x / pracma::trapz(month, fit), .names = "{.col}_norm"))

or_peaks <-  find_peaks(or_grid)
or_peaks$fit_norm <- or_peaks$fit / pracma::trapz(or_grid$month, or_grid$fit)

# Clip grid back to 1-12 for plotting
or_norm_plot <- ggplot(or_grid %>% filter(month >= 1, month <= 12), 
       aes(x = month, y = fit_norm)) +
   geom_ribbon(aes(ymin = lwr_norm, ymax = upr_norm), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(data = first(or_peaks), aes(x = month, y = fit_norm),
             size = 3, shape = 21, fill = "white", stroke = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_bw() +
  labs(x = NULL, 
       y = NULL)

# Clip grid back to 1-12 for plotting
or_plot <- ggplot(or_grid %>% filter(month >= 1, month <= 12), 
       aes(x = month, y = fit)) +
   geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(data = first(or_peaks), aes(x = month, y = fit),
             size = 3, shape = 21, fill = "white", stroke = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_bw() +
  labs(x = NULL, 
       y = "Zc")


plot_grid(or_raw_plot, or_plot, 
          nrow = 2, 
          align = "h")

ggsave("figs/or_zc.svg", width = 4, height = 4)
```

### WA GAM

```{r}
wa_zc <- read.csv(here::here("data/other_sites/WA/WA_zc_digitized.csv"), 
                  col.names = c("month", "weekly_mins")) %>% 
  mutate(month = floor(month) + 1, 
         weekly_mins = pmax(weekly_mins, 0)) %>% 
  group_by(month) %>% 
  summarize(zc_mins = sum(weekly_mins))
  

wa_zc_raw_plot <- ggplot(wa_zc, aes(month, zc_mins)) + 
  geom_col() + 
  geom_smooth() + 
    labs(x = "Month", 
       y = "Zc presence (avg mins)") +
  theme_minimal()

wa_mod <- gam(zc_mins ~ 
                s(month, bs = "cc", k = 12), #adjusted from 12 - make sure is ok
              family = tw(link = "log"),
              data = wa_zc,
              method = "fREML")


wa_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(wa_mod,
                newdata = wa_grid,
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

wa_grid <- wa_grid %>%
  mutate(
    fit = exp(pred$fit),
    lwr = exp(pred$fit - 1.96 * pred$se.fit),
    upr = exp(pred$fit + 1.96 * pred$se.fit)
  ) %>% 
  mutate(across(fit:upr, \(x) x / pracma::trapz(month, fit), .names = "{.col}_norm"))

wa_peaks <-  find_peaks(wa_grid)
wa_peaks$fit_norm <- wa_peaks$fit / pracma::trapz(wa_grid$month, wa_grid$fit)

# Clip grid back to 1-12 for plotting
wa_norm_plot <- ggplot(wa_grid %>% filter(month >= 1, month <= 12), 
       aes(x = month, y = fit_norm)) +
   geom_ribbon(aes(ymin = lwr_norm, ymax = upr_norm), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(data = first(wa_peaks), aes(x = month, y = fit_norm),
             size = 3, shape = 21, fill = "white", stroke = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_bw() +
  labs(x = NULL, 
       y = NULL)

```

### GoA plots

```{r}
goa_zc <- read.csv(here::here("data/other_sites/GoA/Zc_Pratt_GoA.csv"), 
                   col.names = c("month", "weekly_mins")) %>% 
  mutate(month = floor(month) + 1, 
         weekly_mins = pmax(weekly_mins, 0)) %>% 
  group_by(month) %>% 
  summarize(zc_mins = sum(weekly_mins))

goa_zc_raw_plot <- ggplot(goa_zc, aes(month, zc_mins)) + 
  geom_col() + 
  geom_smooth() + 
    labs(x = "Month", 
       y = "Zc presence (avg mins)") +
  theme_minimal()

goa_mod <- gam(zc_mins ~ 
                s(month, bs = "cc", k = 12), #adjusted from 12 - make sure is ok
              family = tw(link = "log"),
              data = goa_zc,
              method = "fREML")

goa_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(goa_mod,
                newdata = goa_grid,
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

goa_grid <- goa_grid %>%
  mutate(
    fit = exp(pred$fit),
    lwr = exp(pred$fit - 1.96 * pred$se.fit),
    upr = exp(pred$fit + 1.96 * pred$se.fit)
  ) %>% 
  mutate(across(fit:upr, \(x) x / pracma::trapz(month, fit), .names = "{.col}_norm"))

goa_peaks <-  find_peaks(goa_grid)
goa_peaks$fit_norm <- goa_peaks$fit / pracma::trapz(goa_grid$month, goa_grid$fit)

# Clip grid back to 1-12 for plotting
goa_norm_plot <- ggplot(goa_grid %>% filter(month >= 1, month <= 12), 
       aes(x = month, y = fit_norm)) +
   geom_ribbon(aes(ymin = lwr_norm, ymax = upr_norm), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(data = last(goa_peaks), aes(x = month, y = fit_norm),
             size = 3, shape = 21, fill = "white", stroke = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_bw() +
  labs(x = NULL, 
       y = NULL)
```

### Circular plots

```{r}
breeding <- data.frame(
  month = 1:12,
  breeding = c(0, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 0)  # Jan-Jun, Oct-Dec
)

# normalize each variable to 0-1
zc_plot_scb <- scb_zc %>% 
  group_by(month) %>%
  summarize(Zc_pres_eff_adj = mean(Zc_pres_eff_adj)) %>% 
  mutate(value = (Zc_pres_eff_adj - min(Zc_pres_eff_adj)) / (max(Zc_pres_eff_adj) - min(Zc_pres_eff_adj)))

#how to put site into this?
scb_circ <- ggplot() +
  geom_rect(data = breeding %>% filter(breeding == 1),
            aes(xmin = month - .5, xmax = month + .5,
                ymin = 0, ymax = 1),
            fill = "lightblue", alpha = 0.5,
            inherit.aes = FALSE) +
  geom_col(data = zc_plot_scb,
           aes(x = month, y = value),
           fill = "darkorange",
           alpha = 0.8,
           width = 0.9) +
  scale_x_continuous(breaks = 1:12, 
                     labels = month.abb) +
  coord_polar(start = -pi/12) +
  ylim(0, 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(color = NULL)

zc_plot_sri <- sri_zc %>% 
  group_by(month) %>%
  summarize(mins_eff_adj = mean(mins_eff_adj)) %>% 
  mutate(value = (mins_eff_adj - min(mins_eff_adj)) / (max(mins_eff_adj) - min(mins_eff_adj)))

sri_circ <- ggplot() +
  geom_rect(data = breeding %>% filter(breeding == 1),
            aes(xmin = month - .5, xmax = month + .5,
                ymin = 0, ymax = 1),
            fill = "lightblue", alpha = 0.5,
            inherit.aes = FALSE) +
  geom_col(data = zc_plot_sri,
           aes(x = month, y = value),
           fill = "darkorange",
           alpha = 0.8,
           width = 0.9) +
  scale_x_continuous(breaks = 1:12, 
                     labels = month.abb) +
  coord_polar(start = -pi/12) +
  ylim(0, 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(color = NULL)

zc_plot_or <- or_zc %>% 
  group_by(month) %>%
  summarize(days_eff_adj = mean(days_eff_adj)) %>% 
  mutate(value = (days_eff_adj - min(days_eff_adj)) / (max(days_eff_adj) - min(days_eff_adj)))

or_circ <- ggplot() +
  geom_rect(data = breeding %>% filter(breeding == 1),
            aes(xmin = month - .5, xmax = month + .5,
                ymin = 0, ymax = 1),
            fill = "lightblue", alpha = 0.5,
            inherit.aes = FALSE) +
  geom_col(data = zc_plot_or,
           aes(x = month, y = value),
           fill = "darkorange",
           alpha = 0.8,
           width = 0.9) +
  scale_x_continuous(breaks = 1:12, 
                     labels = month.abb) +
  coord_polar(start = -pi/12) +
  ylim(0, 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(color = NULL)

mb_circ <- ggplot() +
  geom_rect(data = breeding %>% filter(breeding == 1),
            aes(xmin = month - .5, xmax = month + .5,
                ymin = 0, ymax = 1),
            fill = "lightblue", alpha = 0.5,
            inherit.aes = FALSE) +
  geom_col(data = zc_plot,
           aes(x = month, y = value),
           fill = "darkorange",
           alpha = 0.8,
           width = 0.9) +
  scale_x_continuous(breaks = 1:12, 
                     labels = month.abb) +
  coord_polar(start = -pi/12) +
  ylim(0, 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(color = NULL)

zc_plot_wa <- wa_zc %>% 
  group_by(month) %>%
  summarize(zc_mins = mean(zc_mins)) %>% 
  mutate(value = (zc_mins - min(zc_mins)) / (max(zc_mins) - min(zc_mins)))

wa_circ <- ggplot() +
  geom_rect(data = breeding %>% filter(breeding == 1),
            aes(xmin = month - .5, xmax = month + .5,
                ymin = 0, ymax = 1),
            fill = "lightblue", alpha = 0.5,
            inherit.aes = FALSE) +
  geom_col(data = zc_plot_wa,
           aes(x = month, y = value),
           fill = "darkorange",
           alpha = 0.8,
           width = 0.9) +
  scale_x_continuous(breaks = 1:12, 
                     labels = month.abb) +
  coord_polar(start = -pi/12) +
  ylim(0, 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(color = NULL)

zc_plot_goa <- goa_zc %>% 
  group_by(month) %>%
  summarize(zc_mins = mean(zc_mins)) %>% 
  mutate(value = (zc_mins - min(zc_mins)) / (max(zc_mins) - min(zc_mins)))

goa_circ <- ggplot() +
  geom_rect(data = breeding %>% filter(breeding == 1),
            aes(xmin = month - .5, xmax = month + .5,
                ymin = 0, ymax = 1),
            fill = "lightblue", alpha = 0.5,
            inherit.aes = FALSE) +
  geom_col(data = zc_plot_goa,
           aes(x = month, y = value),
           fill = "darkorange",
           alpha = 0.8,
           width = 0.9) +
  scale_x_continuous(breaks = 1:12, 
                     labels = month.abb) +
  coord_polar(start = -pi/12) +
  ylim(0, 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(color = NULL)

```

## GAMS plotted together:

```{r}
all_gams_zc <- plot_grid(goa_norm_plot, wa_norm_plot, 
          or_norm_plot, zc_mb_norm_plot, 
          sri_norm_plot, scb_norm_plot, 
          nrow = 6)

ggsave("figs/all_zc_gams.svg", all_gams_zc, width = 2.25, height = 6.6)

all_circle_zc <- plot_grid(goa_circ, wa_circ, 
          or_circ, mb_circ, 
          sri_circ, scb_circ, 
          nrow = 6)
# 
# ggsave("figs/or_wa_ak.svg", or_wa_ak, height = 7)
# 
# ca <- plot_grid(mb_circ, sri_circ, scb_circ, 
#           ncol = 1)
#           
# ggsave("figs/ca.svg", ca, height = 7)

```

### Map

```{r}
library(marmap)      # ETOPO1 bathymetry
library(metR)        # geom_contour_fill 
library(terra)
library(tidyterra)
library(rnaturalearth)
library(rnaturalearthdata)

bathy <- getNOAA.bathy(
  lon1 = -145, lon2 = -115, 
  lat1 = 30, lat2 = 60,
  keep = TRUE, 
  path = "data/bathy/"
)
bathy_df <- fortify(bathy)
isobaths <- c(-200, -500, -1000, -2000, -4000) 

land <- ne_countries(scale = "medium", returnclass = "sf")
states <- ne_states(country = "united states of america", returnclass = "sf") 

#oregon site will need to be updated, only including one point for socal for now 
sites <- data.frame(
  name = c("San Diego", "Los Angeles", 
           "Monterey", "Oregon", 
           "Washington", "Gulf of Alaska"), 
  lon = c(-119.176167, -120.259167, 
          -122.186833, -125.587333, 
          -125.333333, -142.616667), 
  lat = c(32.846000, 33.540750,
          36.712500, 44.656000, 
          47.500000, 56.216667)
)

bw_graticule <- graticule(
  lon = seq(-145, -115, by = 5),
  lat = seq(30, 60, by = 5),
  crs = "EPSG:4326"
)
  
bw_map <- ggplot() + 
  #bathymetry
  geom_raster(data = filter(bathy_df, z <= 0), 
              aes(x, y, fill = z), 
              interpolate = TRUE) + 
  geom_sf(data = bw_graticule,
          color = "grey90",
          alpha = 0.2,
          linewidth = 0.5) +
  scale_fill_gradientn(
    colours  = c("#08306b", "#2171b5", "#6baed6", "#c6dbef"),
    values   = scales::rescale(c(-6000, -2000, -200, 0)),
    name     = "Depth (m)",
    na.value = "grey80"
  ) +
  # Land + state borders
  geom_sf(data = land,
          fill = "grey75",
          colour = NA) +
  geom_sf(data = states, 
          fill = NA, 
          colour = "grey60", 
          linewidth = 0.2) +  
  coord_sf(
    xlim   = c(-144, -117.5),
    ylim   = c(31, 60),
    expand = FALSE
  ) + 
  scale_x_continuous(breaks = seq(-145, -115, by = 5)) +
  # Isobath contours
  geom_contour(
    data    = bathy_df,
    aes(x, y, z = z),
    breaks  = isobaths,
    colour  = "lightblue", linewidth = 0.25, alpha = 0.6
  ) + 
  geom_point(data = sites, 
             aes(lon, lat), 
             color = "white", 
             size = 2, 
             inherit.aes = FALSE) +
  geom_hline(yintercept = c(32.8, 33.5, 
                            36.67, 44.67, 
                            47.5, 56.25), 
             color = "white", 
             linetype = "dashed") + 
  annotate("text", x = -137, y = c(33.1, 33.75, 
                                     37, 45, 
                                     47.8, 56.5),
         label = c("32.8°N", "33.5°N",
                   "36.6°N", "44.7°N", 
                   "47.5°N", "56°N"),
         colour = "white", size = 2.5, hjust = 1) + 
  geom_text(data = sites,
            aes(x = lon, y = lat, label = name),
            fontface = "bold",
            color = "white",
            nudge_y = 0.4, 
            nudge_x = 0.5, 
            size = 2.5) +
  # labs(x = "Latitude", y = "Longitude") + 
  theme_minimal(12) + 
  theme(legend.position = "inside",
        legend.position.inside = c(0.98, 0.98),
        legend.justification = c(1, 1), 
        legend.background = element_rect(fill = "white"))

bw_map
ggsave("figs/zc_map.svg", bw_map, width = 4, height = 8)
```

```         
```

## Oregon lag analysis

```{r}
Xp_zc_or <- predict(or_mod, 
                 newdata = or_grid, 
                 type = "lpmatrix")
coef_sims_zc_or <- MASS::mvrnorm(n_sim, 
                              mu = coef(or_mod), 
                              Sigma = vcov(or_mod))
sim_curves_zc_or <- Xp_zc_or %*% t(coef_sims_zc_or)
sim_curves_response_zc_or <- or_mod$family$linkinv(sim_curves_zc_or)
boot_zc_or <- expand_grid(i = 1:nrow(or_grid), 
            j = 1:ncol(sim_curves_response_zc_or)) %>% 
  mutate(month = or_grid$month[i], 
         sim = j, 
         count = sim_curves_response_zc_or[cbind(i, j)]) %>% 
  select(-c(i, j)) 

##beuti setup for oregon##
beuti_or <- read_csv(here::here("data/indices/BEUTI_06-25.csv")) %>% 
  set_names(c("date", "latitude", "beuti")) %>% 
  filter(latitude == 45) %>% 
  mutate(month = month(date), 
         year = year(date))

#climatology to get at the long term average
beuti_clim_mod_or <- gam(beuti ~ 
                        s(month, bs = "cc", k = 12) +
                        s(year, bs = "re"),
                      family = gaussian(),
                      data = beuti_or,
                      method = "REML")

or_beuti_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(beuti_clim_mod_or,
                newdata = or_beuti_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

or_beuti_grid$beuti <- pred$fit
or_beuti_grid$lower <- pred$fit - pred$se.fit * 1.96
or_beuti_grid$upper <- pred$fit + pred$se.fit * 1.96

or_beuti_grid %>% 
  filter(between(month, 1, 12)) %>% 
  ggplot(aes(month, beuti)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line() + 
  theme_classic() +
  scale_x_continuous(breaks = 1:12) + 
  geom_vline(xintercept = or_peaks$month, 
             color = "tomato", 
             linetype = "dashed") + 
  ggtitle("Oregon")

ggsave("figs/or_beuti.svg", width = 2.8, height = 2.5)

# Peak lasts 2.3 months
sum(or_beuti_grid$beuti > 0.9 * max(or_beuti_grid$beuti)) * 0.1
or_beuti_grid$month[which.max(or_beuti_grid$beuti)]


### BEUTI curve sim ###
Xp_beuti_or <- predict(beuti_clim_mod_or, newdata = or_beuti_grid, type = "lpmatrix")
coef_sims_beuti_or <- MASS::mvrnorm(n_sim, mu = coef(beuti_clim_mod_or), Sigma = vcov(beuti_clim_mod_or))
sim_curves_beuti_or <- Xp_beuti_or %*% t(coef_sims_beuti_or)

#calculating lags
boot_beuti_or <- expand_grid(i = 1:nrow(or_beuti_grid), 
            j = 1:ncol(sim_curves_beuti_or)) %>% 
  mutate(month = or_beuti_grid$month[i], 
         sim = j, 
         beuti = sim_curves_beuti_or[cbind(i, j)]) %>% 
  select(-c(i, j)) 

boot_beuti_peak_or <- boot_beuti_or %>% 
  group_by(sim) %>% 
  summarize(peak_month_beuti = month[which.max(beuti)])

boot_zc_peak_or <- boot_zc_or %>% 
  group_by(sim) %>% 
  summarize(peak_month_zc = month[which.max(count)])

boot_peak_lag_zc_or <- left_join(boot_beuti_peak_or, boot_zc_peak_or, by = "sim") %>% 
  mutate(peak_lag = peak_month_zc - peak_month_beuti, 
         peak_lag = ifelse(peak_lag <= 0, peak_lag + 12, peak_lag)) 

mean_peak_lag_zc_or <- CircStats::circ.mean(boot_peak_lag_zc_or$peak_lag * 2 * pi / 12) * 12 / 2 / pi
ci_peak_lag_zc_or <- shortest_arc_ci(boot_peak_lag_zc_or$peak_lag)

if (mean_peak_lag_zc_or < 0) mean_peak_lag_zc_or <- mean_peak_lag_zc_or + 12

ggplot(boot_peak_lag_zc_or, aes(peak_lag)) + 
  geom_histogram() + 
  geom_vline(xintercept = mean_peak_lag_zc_or, color = "red") + 
  ggtitle("Zc") + 
  theme_classic()

ggsave("figs/or_lag.svg", width = 2.8, height = 2.5)

```

Or circle plot

```{r}
# normalize each variable to 0-1
beuti_clim_or <- beuti_or %>% 
  group_by(latitude, month) %>% 
  summarize(beuti = mean(beuti))

beuti_plot_or <- beuti_clim_or %>% 
  mutate(value = (beuti - min(beuti)) / (max(beuti) - min(beuti)))

or_circ_large <- ggplot() +
    geom_rect(data = brooding %>% filter(brooding == 1),
            aes(xmin = month - .5, xmax = month + .5,
                ymin = 0, ymax = 1),
            fill = "darkred", alpha = 0.2,
            inherit.aes = FALSE) +
  geom_col(data = zc_plot_or,
           aes(x = month, y = value),
           fill = "darkorange",
           alpha = 0.8,
           width = 0.9) +
  scale_x_continuous(breaks = 1:12, 
                     labels = month.abb) +
  geom_line(data = beuti_plot_or, 
            aes(x = month, y = value), 
            color = "steelblue", 
            linewidth = 1) +
  coord_polar(start = -pi/12) +
  ylim(0, 1) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(color = NULL)

ggsave("figs/or_circle.svg", width = 2.8, height = 2.5)

```
