---
title: "glowingwave"
format: html
---

Do goosebeaked whales travel up and down the US West Coast in a seasonal pattern?

### BW Setup 

from three sites: Southern CA Bight, Monterey Bay, and Newport, OR.

```{r}
#| message: false
#| label: setup

library(tidyverse)
library(lubridate)
library(cowplot)
library(mgcv)
source(here::here("R/bw_funcs.R"))

#southern california bight
scb_bw <- rbind(
  read_csv("~/Downloads/SiteN_Zc_monthly_bin.csv") %>% 
    mutate(site = "N"),
  read_csv("~/Downloads/SiteH_Zc_monthly_bin.csv") %>% 
    mutate(site = "H")
) %>% 
  mutate(date = dmy(bin),
         month = month(date),
         year = year(date),
         site = factor(site)) %>% 
  arrange(site, date) %>% 
  drop_na(Zc_pres_eff_adj) %>% 
  group_by(site)

#monterey
monterey_bw <- read_csv(here::here("data/zc_monterey.csv")) %>% 
  mutate(date = ymd(Date), 
         month = Month,
         year = year(Date)) %>% 
  arrange(date) %>% 
  complete(year = 2017:2018, month = 1:12, fill = list(count = 0)) %>% 
  distinct(year, month, count) %>% 
  mutate(date = ISOdate(year, month, day = 15))

#oregon
oregon_bw <- read_csv(here::here("data/othersites/monthly_presence_zc_oregon.csv")) %>% 
  mutate(date = ymd(month_year), 
         month = month(month_year),
         year = year(month_year)) %>% 
  arrange(date) %>% 
  complete(year = 2022, month = 1:12, fill = list(days_present = 0))
```

### Raw detections at each site. 

```{r}
#| label: raw-bw-data-plot
#| fig-height: 12

#scb
scb_p1 <- ggplot(scb_bw, aes(factor(month), Zc_pres_eff_adj, color = site)) +
  geom_boxplot() +
  labs(x = "Month", 
       y = "Zc presence (minutes)", 
       title = "SCB") +
  theme_bw() +
  theme(legend.position = "top") 

#monterey
mb_p1 <- ggplot(monterey_bw, aes(factor(month), count)) +
  geom_point(aes(color = factor(year))) +
  labs(x = "Month", 
       y = "Zc presence (days)", 
       title = "MB") +
  theme_bw()

#oregon
or_p1 <- ggplot(oregon_bw, aes(factor(month), days_present)) +
  geom_point() +
  labs(x = "Month", 
       y = "Zc presence (days)", 
       title = "OR") +
  theme_bw()

plot_grid(scb_p1, mb_p1, or_p1, nrow = 3, align = "h")
```

Fitting GAMS to show seasonal presence at each site.

### SCB BW GAM

```{r}
#| label: scripps-gam

scb_bw_mod <- gam(Zc_pres_eff_adj ~ 
                    s(month, by = site, bs = "cc", k = 12) +
                    site +
                    s(year, bs = "re"),
                  family = tw(link = "log"),
                  data = scb_bw,
                  method = "fREML",
                  knots = list(month = c(1, 12)))

summary(scb_bw_mod)
gam.check(scb_bw_mod)
acf(residuals(scb_bw_mod))

# Residuals are autocorrelated, but for peak timing we don't care. Be careful about p-values etc.

scb_bw_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  site  = unique(scb_bw$site),
  year  = 0
) %>% mutate(site = factor(site, levels = levels(scb_bw$site)))

pred <- predict(scb_bw_mod,
                newdata = scb_bw_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

scb_bw_grid <- scb_bw_grid %>%
  mutate(
    fit = exp(pred$fit),
    lwr = exp(pred$fit - 1.96 * pred$se.fit),
    upr = exp(pred$fit + 1.96 * pred$se.fit)
  ) %>% 
  mutate(across(fit:upr, \(x) x / pracma::trapz(month, fit), .names = "{.col}_norm"))

# Find peaks on extended grid, then fold month back to 1-12
peaks <- scb_bw_grid %>%
  group_by(site) %>%
  group_modify(~ find_peaks(.x)) %>%
  ungroup()

peaks$fit_norm <- peaks$fit / pracma::trapz(scb_bw_grid$month, scb_bw_grid$fit)

# Clip grid back to 1-12 for plotting
scb_norm_plot <- ggplot(scb_bw_grid %>% filter(month >= 1, month <= 12), 
       aes(x = month, y = fit_norm, color = site)) +
  geom_ribbon(aes(ymin = lwr_norm, ymax = upr_norm, fill = site), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(data = peaks, aes(x = month, y = fit_norm, color = site),
             size = 3, shape = 21, fill = "white", stroke = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  facet_wrap(~ site, ncol = 1, scales = "free_y") +
  theme_bw() +
  theme(legend.position = "top") +
  labs(x = NULL, y = "Zc presence", color = "Site", fill = "Site", 
       title = "Southern California Bight | 33° N", 
       subtitle = "2007-2020") 
```

### MB BW GAM

```{r}
#| label: mb-gam

monterey_bw_mod <- gam(count ~ 
                       s(month, bs = "cc", k = 12) +
                       s(year, bs = "re"),
                     family = tw(link = "log"),
                     data = monterey_bw,
                     method = "fREML")

summary(monterey_bw_mod)
gam.check(monterey_bw_mod)
acf(residuals(monterey_bw_mod))

monterey_bw_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(monterey_bw_mod,
                newdata = monterey_bw_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

monterey_bw_grid <- monterey_bw_grid %>%
  mutate(
    fit = exp(pred$fit),
    lwr = exp(pred$fit - 1.96 * pred$se.fit),
    upr = exp(pred$fit + 1.96 * pred$se.fit)
  ) %>% 
  mutate(across(fit:upr, \(x) x / pracma::trapz(month, fit), .names = "{.col}_norm"))

peaks_mb <-  find_peaks(monterey_bw_grid)
peaks_mb$fit_norm <- peaks_mb$fit / pracma::trapz(monterey_bw_grid$month, monterey_bw_grid$fit)

# Clip grid back to 1-12 for plotting
mb_norm_plot <- ggplot(monterey_bw_grid %>% filter(month >= 1, month <= 12), 
       aes(x = month, y = fit_norm)) +
  # geom_ribbon(aes(ymin = lwr_norm, ymax = upr_norm), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(data = peaks_mb[2,], aes(x = month, y = fit_norm),
             size = 3, shape = 21, fill = "white", stroke = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_bw() +
  labs(x = NULL, 
       y = "Zc presence", 
       title = "Monterey Bay, California | 36° N", 
       subtitle = "2017-2018") 
```

### OR BW GAM

```{r}
#| label: or-gam

oregon_bw_mod <- gam(days_present ~ 
                    s(month, bs = "cc", k = 12) +
                    s(year, bs = "re"),
                  family = tw(link = "log"),
                  data = oregon_bw,
                  method = "fREML")

summary(oregon_bw_mod)
gam.check(oregon_bw_mod)
acf(residuals(oregon_bw_mod))

oregon_bw_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(oregon_bw_mod,
                newdata = oregon_bw_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

oregon_bw_grid <- oregon_bw_grid %>%
  mutate(
    fit = exp(pred$fit),
    lwr = exp(pred$fit - 1.96 * pred$se.fit),
    upr = exp(pred$fit + 1.96 * pred$se.fit)
  ) %>% 
  mutate(across(fit:upr, \(x) x / pracma::trapz(month, fit), .names = "{.col}_norm"))

or_peaks <-  find_peaks(oregon_bw_grid)
or_peaks$fit_norm <- or_peaks$fit / pracma::trapz(oregon_bw_grid$month, oregon_bw_grid$fit)

# Clip grid back to 1-12 for plotting
or_norm_plot <- ggplot(oregon_bw_grid %>% filter(month >= 1, month <= 12), 
       aes(x = month, y = fit_norm)) +
  # geom_ribbon(aes(ymin = lwr_norm, ymax = upr_norm), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(data = first(or_peaks), aes(x = month, y = fit_norm),
             size = 3, shape = 21, fill = "white", stroke = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_bw() +
  labs(x = NULL, 
       y = "Zc presence", 
       title = "Newport, Oregon | 45° N", 
       subtitle = "2021-2022") 

```

### Wave plot

```{r}
#| label: wave-plot
#| fig-height: 12


plot_grid(or_norm_plot, mb_norm_plot, scb_norm_plot, nrow = 3, rel_heights = c(1, 1, 2))
```

Now we want to see if there is a relationship between Zc presence and environmental conditions. We'll use BEUTI, an index based on biological upwelling metrics.

### BEUTI setup

```{r}
#| label: beuti-setup

beuti <- read_csv(here::here("data/indices/BEUTI_06-25.csv")) %>% 
  set_names(c("date", "latitude", "beuti")) %>% 
  filter(latitude %in% c(33, 36, 45)) %>% 
  mutate(month = month(date), 
         year = year(date))

scb_beuti <- filter(beuti, latitude == 33)
mb_beuti <- filter(beuti, latitude == 36)
or_beuti <- filter(beuti, latitude == 45)

#climatology - gets at the long term average
beuti_clim <- beuti %>% 
  group_by(latitude, month) %>% 
  summarize(beuti = mean(beuti))

beuti_raw_plot <- beuti_clim %>% 
  mutate(latitude = factor(latitude)) %>% 
  ggplot(aes(month, 
             beuti, 
             color = latitude)) + 
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "BEUTI") +
  theme_bw() + 
  theme(legend.position = c(0.98, 0.98),
        legend.justification = c(1, 1), 
        legend.background = element_rect(fill = "transparent"))

beuti_clim_mod <- gam(beuti ~ 
                        s(month, bs = "cc", k = 12) +
                        s(year, bs = "re"),
                      family = gaussian(),
                      data = mb_beuti,
                      method = "REML")
summary(beuti_clim_mod)
```

### MB BEUTI GAM

```{r}
mb_beuti_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(beuti_clim_mod,
                newdata = mb_beuti_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

mb_beuti_grid$beuti <- pred$fit
mb_beuti_grid$lower <- pred$fit - pred$se.fit * 1.96
mb_beuti_grid$upper <- pred$fit + pred$se.fit * 1.96

mb_beuti_grid %>% 
  filter(between(month, 1, 12)) %>% 
  ggplot(aes(month, beuti)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line() + 
  theme_classic() +
  scale_x_continuous(breaks = 1:12) + 
  ylim(-5, 20) + 
  ggtitle("Monterey Bay")
```

### OR BEUTI GAM

```{r}
#fitting the gam for oregon beuti
or_beuti_clim_mod <- gam(beuti ~ 
                        s(month, bs = "cc", k = 12) +
                        s(year, bs = "re"),
                      family = gaussian(),
                      data = or_beuti,
                      method = "REML")
summary(or_beuti_clim_mod)

or_beuti_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(or_beuti_clim_mod,
                newdata = or_beuti_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

or_beuti_grid$beuti <- pred$fit
or_beuti_grid$lower <- pred$fit - pred$se.fit * 1.96
or_beuti_grid$upper <- pred$fit + pred$se.fit * 1.96

or_beuti_grid %>% 
  filter(between(month, 1, 12)) %>% 
  ggplot(aes(month, beuti)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line() + 
  theme_classic() +
  scale_x_continuous(breaks = 1:12) + 
  ylim(-5, 20) + 
  ggtitle("Oregon")
```

### SCB BEUTI GAM

```{r}
#fitting the gam for oregon beuti
scb_beuti_clim_mod <- gam(beuti ~ 
                        s(month, bs = "cc", k = 12) +
                        s(year, bs = "re"),
                      family = gaussian(),
                      data = scb_beuti,
                      method = "REML")

scb_beuti_grid <- expand.grid(
  month = seq(0, 13, by = 0.1),   # extend beyond 1-12
  year  = 0)

pred <- predict(scb_beuti_clim_mod,
                newdata = scb_beuti_grid,
                exclude = "s(year)",
                newdata.guaranteed = TRUE,
                type = "link",
                se.fit = TRUE)

scb_beuti_grid$beuti <- pred$fit
scb_beuti_grid$lower <- pred$fit - pred$se.fit * 1.96
scb_beuti_grid$upper <- pred$fit + pred$se.fit * 1.96

scb_beuti_grid %>% 
  filter(between(month, 1, 12)) %>% 
  ggplot(aes(month, beuti)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  geom_line() + 
  theme_classic() +
  scale_x_continuous(breaks = 1:12) + 
  ylim(-5, 20) + 
  ggtitle("Southern California")
```

simulate curves for that bad boy

```{r}
n_sim <- 100
Xp <- predict(beuti_clim_mod, newdata = mb_beuti_grid, type = "lpmatrix")
coef_sims <- MASS::mvrnorm(n_sim, mu = coef(beuti_clim_mod), Sigma = vcov(beuti_clim_mod))
sim_curves <- Xp %*% t(coef_sims)
#will need to transform scale for beaked whale one: sim_curves_response <- gam_mod$family$linkinv(sim_curves)
expand_grid(i = 1:nrow(mb_beuti_grid), 
            j = 1:ncol(sim_curves)) %>% 
  mutate(month = mb_beuti_grid$month[i], 
         sim = j, 
         beuti = sim_curves[cbind(i, j)]) %>% 
  select(-c(i, j)) %>% 
  ggplot(aes(month, beuti, group = sim)) + 
  geom_line(alpha = 0.1)

```

```{r}

scb_beuti_bw <- expand_grid(beuti, site = c("H", "N")) %>%
  select(site, month, year, beuti) %>% 
  mutate(site = as.character(site)) %>% 
  full_join(scb_bw %>% mutate(site = as.character(site)),
            by = c("month", "year", "site")) %>% 
  arrange(year, month, site) %>% 
  mutate(site = factor(site))

scb_beuti_mod <- gam(Zc_pres_eff_adj ~ 
                    s(beuti, by = site) +
                    site +
                    s(year, bs = "re"),
                  family = tw(link = "log"),
                  data = scb_beuti_bw,
                  method = "fREML")

scb_beuti_plot <- scb_beuti_bw %>% 
  group_by(site) %>% 
  # slice(-(1:max_lag)) %>% 
  mutate(across(c(beuti, Zc_pres_eff_adj), 
                \(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))) %>% 
  ungroup() %>% 
  pivot_longer(c(beuti, Zc_pres_eff_adj), 
               names_to = "var", 
               values_to = "val") %>% 
  ggplot(aes(date, val, color = var)) +
  geom_line() +
  facet_grid(rows = vars(site)) +
  ggtitle("Southern California Bight (33°N)") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.title = element_blank(), 
        plot.margin = margin(t = 15, r = 5, b = 15, l = 5)) 
```

```{r}
max_lag <- 12
lag_mods <- map(0:max_lag, \(.lag) {
  lagged_beuti <- scb_beuti_bw %>% 
    group_by(site) %>% 
    mutate(beuti = lag(beuti, .lag)) %>% 
    slice(-c(1:max_lag)) %>% 
    ungroup()
  gam(Zc_pres_eff_adj ~ 
        s(beuti, by = site) +
        site +
        s(year, bs = "re"),
      family = tw(link = "log"),
      data = lagged_beuti,
      method = "fREML")
})
lag_aic <- map_dbl(lag_mods, AIC)
tibble(lag = 0:max_lag,
       aic = lag_aic,
       daic = lag_aic - min(lag_aic))

#0 lag is best for socal
```

```{r}
mb_beuti_bw <- beuti_clim %>%
  filter(latitude == 36) %>% 
  full_join(monterey_bw,
            by = c("month")) 

mb_beuti_mod <- gam(count ~ 
                    s(beuti) +
                    s(year, bs = "re"),
                  family = tw(link = "log"),
                  data = mb_beuti_bw,
                  method = "fREML")

mb_beuti_plot <- mb_beuti_bw %>% 
  # slice(-(1:max_lag)) %>% 
  mutate(across(c(beuti, count), 
                \(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))) %>% 
  ungroup() %>% 
  pivot_longer(c(beuti, count), 
               names_to = "var", 
               values_to = "val") %>% 
  ggplot(aes(date, val, color = var)) +
  geom_line() +
  ggtitle("Monterey (36°N)") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.title = element_blank(), 
        plot.margin = margin(t = 15, r = 5, b = 15, l = 5)) 
```

```{r}
lag_mods_mb <- map(0:max_lag, \(.lag) {
  mb_lagged_beuti <- mb_beuti_bw %>% 
    mutate(beuti = lag(beuti, .lag)) %>% 
    slice(-c(1:max_lag))
  gam(count ~ 
        s(beuti, k = 5) +
        s(year, bs = "re"),
      family = tw(link = "log"),
      data = mb_lagged_beuti,
      method = "fREML")
})
mb_lag_aic <- map_dbl(lag_mods_mb, AIC)
tibble(lag = 0:max_lag,
       aic = mb_lag_aic,
       daic = mb_lag_aic - min(mb_lag_aic)) 

#best lag for this is 6 months

```

```{r}
or_beuti_bw <- beuti_clim %>%
  filter(latitude == 45) %>% 
  full_join(oregon_bw,
            by = c("month")) 

or_beuti_mod <- gam(days_present ~ 
                    s(beuti) +
                    s(year, bs = "re"),
                  family = tw(link = "log"),
                  data = or_beuti_bw,
                  method = "fREML")

or_beuti_plot <- or_beuti_bw %>% 
  # slice(-(1:max_lag)) %>% 
  mutate(across(c(beuti, days_present), 
                \(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))) %>% 
  ungroup() %>% 
  pivot_longer(c(beuti, days_present), 
               names_to = "var", 
               values_to = "val") %>% 
  ggplot(aes(date, val, color = var)) +
  geom_line() +
  ggtitle("Oregon (45°N)") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.title = element_blank(), 
        plot.margin = margin(t = 15, r = 5, b = 15, l = 5))
```

```{r}

lag_mods_or <- map(0:max_lag, \(.lag) {
  or_lagged_beuti <- or_beuti_bw %>% 
    mutate(beuti = lag(beuti, .lag)) %>% 
    slice(-c(1:max_lag))
  #not enough data for a spline currently - add back in with more years
  gam(days_present ~ 
        beuti +
        s(year, bs = "re"),
      family = tw(link = "log"),
      data = or_lagged_beuti,
      method = "fREML")
})
or_lag_aic <- map_dbl(lag_mods_or, AIC)
tibble(lag = 0:max_lag,
       aic = or_lag_aic,
       daic = or_lag_aic - min(or_lag_aic)) 

#best lag for this is 7 months
```

```{r}
site_beuti_plots <- cowplot::plot_grid(scb_beuti_plot, mb_beuti_plot, or_beuti_plot, 
                                       ncol = 1, rel_heights = c(2, 1, 1))
cowplot::plot_grid(beuti_raw_plot, site_beuti_plots,
                   ncol = 2, rel_widths = c(1, 1))
```

```{r}
#pretty sure this is not going to work
# hci <- read_csv("data/indices/HCI.csv") %>% 
#   mutate(time = lubridate::ym(time))
# 
# 
# mb_hci <- hci %>% 
#   filter(timeseries == "Monthly Habitat Compression Index over 35.5-40N")   
# 
# mb_hci %>% 
#   filter(year > 2010) %>% 
#   ggplot(aes(time, index, group = 1)) + 
#   geom_line()
```

```{r}
# beuti <- read_csv(here::here("data/indices/BEUTI_06-25.csv")) %>% 
#   set_names(c("date", "latitude", "beuti")) %>% 
#   filter(latitude %in% c(33, 36, 45)) %>% 
#   mutate(month = month(date), 
#          year = year(date))
# 
# #climatology - gets at the long term average
# beuti_clim <- beuti %>% 
#   group_by(latitude, month) %>% 
#   summarize(beuti = mean(beuti))
# 
# beuti_clim %>% 
#   mutate(latitude = factor(latitude)) %>% 
#   ggplot(aes(month, 
#              beuti, 
#              color = latitude)) + 
#   geom_point() + 
#   geom_line() +
#   theme_bw()

```

Summary stats

```{r}
monterey_bw %>% 
  summarize(mean(count), 
            sd(count))

oregon_bw %>% 
  summarize(mean(days_present), 
            sd(days_present))

scb_bw %>% 
  group_by(site) %>% 
  summarize(mean(Zc_pres_eff_adj), 
            sd(Zc_pres_eff_adj))
```
